Parsing Postfix log files

### 2008-02-25 doc/logparser.tex
Updated 'aborted delivery attempts' section: there's no longer a restriction on     the number of smtpd log lines. Documented out of order bounce line complication.

\newcommand{\numberOFrules}[0]{%
    146%
}

@@

    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) logs the
        creation of bounce messages.  This action creates a new mail if
        necessary; if the mail already exists the unknown origin flag will
        be removed.  The action also marks the mail as a bounce
        notification.

@@

Once again Postfix does not log a message making the client's behaviour
clear, so once again heuristics are required to identify when this
behaviour occurs.  In this case a list of all mails accepted during a
connection is saved in the connection state, and the accepted mails are
examined when the disconnection action is executed.  Each mail is checked
for the following characteristics:

    \item Were there exactly two \daemon{smtpd} log lines for the mail?
        There should be a connection line and a mail accepted line, nothing
        else.

If both checks are successful then the mail is assumed to be one of the
offending bogus mails and is discarded.  There will be no further entries
logged for such mails, so without identifying and discarding them they
accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and \IP{} address, but the database schema
requires many more fields be populated (see \sectionref{connections table}
and \sectionref{results table}).

@@

>>>>

\newcommand{\numberOFrules}[0]{%
    157%
}

@@

    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) logs the
        creation of bounce messages.  This action creates a new mail if
        necessary; if the mail already exists the unknown origin flag will
        be removed.  The action also marks the mail as a bounce
        notification.  To deal with complication \sectionref{Further out of
        order log lines} this action checks a cache of recent bounce mails
        to avoid creating bogus bounce mails when lines are logged out of
        order.

@@

Once again Postfix does not log a message making the client's behaviour
clear, so once again heuristics are required to identify when this
behaviour occurs.  In this case a list of all mails accepted during a
connection is saved in the connection state, and the accepted mails are
examined when the disconnection action is executed.  Each mail is checked
for the following: is the mail missing its \daemon{cleanup} log message?
Every mail which passes through Postfix will have a \daemon{cleanup} line;
lack of a \daemon{cleanup} line is a sure sign the mail didn't make it too
far.

If this check is successful then the mail is assumed to be one of the
offending bogus mails and is discarded.  There will be no further entries
logged for such mails, so without identifying and discarding them they
accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and \IP{} address, but the database schema
requires many more fields be populated (see \sectionref{connections table}
and \sectionref{results table}).

@@

\subsubsection{Further out of order log lines}

\label{Further out of order log lines}

This is yet another complication which only occurs during periods of
extremely high load, when process scheduling vagaries and even hard disk
access times cause strange behaviour.  In this complication bounce
notification mails are created, delivered and deleted from the queue,
\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
with this the COMMIT action maintains a cache of recently committed bounce
notification mails, which the BOUNCE action subsequently checks if the
bounce mail isn't already in the state tables. If a mail exists in the
cache under the correct queueid, and its start time is less than ten
seconds before the timestamp of the bounce log line, it is assumed that the
bounce notification mail has already been processed and the BOUNCE action
does not create one.  If the queueid exists it is removed from the cache,
as it has either just been used or it is too old to use in future.  Whether
the BOUNCE action creates a mail or finds one existing in the state tables,
it flags the mail as having been seen by the BOUNCE action; if this flag is
present the COMMIT action will not add the mail to the cache of recent
bounce notification mails.\footnote{This is not required to correctly deal
with the complication, but is an optimisation to reduce memory usage of the
parser; on the occasions the author has observed this action occurring
there have been a huge number of bounce notification mails generated; if
every bounce notification mail was cached it would dramatically increase
the memory requirements of the parser.}  The cache of bounce notification
mails will be pruned whenever the parser's state is saved, though if the
size of the cache ever becomes a problem it could be pruned periodically to
keep it in check.
